### 2022-01-20 README.Rmd
first full draft!

---
title: "Longer term All-Cause and Cardiovascular Mortality with Intensive Blood Pressure Control: A Secondary Analysis of SPRINT"
bibliography: refs.bib
csl: jama.csl
always_allow_html: true
output: 
  officedown::rdocx_document:
    reference_docx: style_manuscript_times_new_roman.docx
---
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE,
                      warning = FALSE,
                      dpi = 600, 
                      fig.width = 11, 
                      fig.height = 6.25)

# specification for table tables
rspec <- round_spec() |>
  round_using_magnitude(digits = c(2, 1, 0),
                        breaks = c(10, 100, Inf))

@@

# default is double space 
# single spacing for title page, abstract, captions
single_par <- fp_par(line_spacing=1)

@@

Byron C. Jaeger, PhD,^1^ Adam P. Bress, PharmD, MS,^2,3^ Joshua D. Bundy, PhD,^4^ Alfred K. Cheung, MD,^5,6^ William C. Cushman, MD,^7^ Paul E. Drawz, MD,^8^ Lawrence J. Fine, MD,^9^ Karen C. Johnson, MD,^7^ Cora E. Lewis, MD,^10^ Suzanne Oparil, MD,^11^ Michael V. Rocco, MD,^12^ Stephen R. Rapp, PhD, ^13^ Joni K. Snyder, RN,BSN,^9^ Mark A. Supiano, MD,^14^ Paul K. Whelton, MD,^4^ Jeff D. Williamson, MD, MHS,^15^ Jackson T. Wright, Jr, MD, PhD,^16^ David M. Reboussin, PhD,^1^ and Nicholas M. Pajewski, PhD^1^

^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC. ^2^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT. ^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT. ^4^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA. ^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT. ^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT. ^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis. ^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN. ^9^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD. ^10^Department of Epidemiology, University of Alabama at Birmingham. ^11^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham. ^12^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC. ^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wakfe Forest School of Medicine, Winston-Salem, NC. ^14^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City. ^15^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC. ^16^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
\

**Correspondence to:** \
Nicholas M. Pajewski, PhD \
Department of Biostatistics and Data Science \
Division of Public Health Sciences \
Wake Forest School of Medicine \
Medical Center Boulevard \
Winston-Salem, NC 27154 \
(336) 713-1396 \
npajewsk@wakehealth.edu `r single_par`

**Word Count:** Abstract: ; Manuscript:

@@

__IMPORTANCE__. Both the Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) trial have shown benefits of intensive blood pressure control on cardiovascular morbidity and mortality. However, as both trials were ended after slightly more than 3 years of follow-up, the effect of intensive treatment on longer term mortality is unknown.

__OBJECTIVE__. To evaluate the legacy effect of intensive hypertension treatment on longer term all-cause and cardiovascular mortality.

__DESIGN, SETTING, AND PARTICIPANTS__. SPRINT, a randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke. Randomization began on November 8, 2010, the trial intervention was stopped early on August 20, 2015, and trial follow-up visits occurred through July 2016.

__INTERVENTIONS__. Randomization to a systolic blood pressure goal of either less than 120 mm Hg (intensive treatment, n=4678) versus less than 140 mm Hg (standard treatment, N=4683).

__MAIN OUTCOMES AND MEASURES__. Cardiovascular and all-cause mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020. Outpatient blood pressures measured in routine clinical practice after the trial were examined in a subset of trial participants (N=3644).

__RESULTS__. Among 9361 randomized participants (mean age, 67.9 years; 3332 women [35.6%]), the median intervention period was 3.34 years. Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial. However, there was no indication of benefit during post-trial follow-up for either cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`). Results were similar for subgroups based on baseline age, cognitive function, and frailty status. Analyses of outpatient blood pressures indicated a steady decline in the mean between group difference following the trial, largely driven by increases in mean systolic blood pressure in participants randomized to intensive treatment, increasing from a mean of 131.7 mm Hg (95% CI 130.7 to 132.6 mm Hg) at 4.5 years of follow-up to 136.8 mm Hg (95% CI 135.7 to 137.8 mm Hg) at 9 years of follow-up.

__CONCLUSIONS AND RELEVANCE__. The observed benefit of intensive treatment on cardiovascular and all-cause mortality was largely attenuated during post-trial observational follow-up. Given increasing blood pressures in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.

# INTRODUCTION

Hypertension is the leading modifiable risk factor for cardiovascular disease (CVD).[@gdbJACC2020] A meta analysis of pooled individual-level data from randomized trials has shown that pharmacological blood pressure (BP) reduction reduces the risk of major cardiovascular events across the spectrum of initial BP, with larger absolute risk reduction for older adults.[@rahimi_age_stratified_2021] The Systolic Blood Pressure Intervention Trial (SPRINT) and the Strategy of Blood Pressure Intervention in the Elderly Hypertensive Patients (STEP) each showed that intensive systolic BP (SBP) control, defined by a SBP target < 120 mm Hg in SPRINT and < 130 mm Hg in STEP, reduced the risk for cardiovascular morbidity and mortality. [@sprint_research_group_randomized_2015; @zhang_trial_2021] However,
only SPRINT showed a benefit of intensive treatment on all-cause mortality. Another contrasting result was that intensive treatment in SPRINT, but not STEP, increased the risk of acute kidney injury (AKI). While this may be attributable to the much lower prevalence of chronic kidney disease in STEP (2.3% vs 28.3%), the consistent association of AKI with increased mortality risk[@chawla2014aki; @rocco2018aki] raises questions about the longer term effect of intensive treatment on CVD and all-cause mortality.

Because both trials were stopped early after a median follow-up of 3 years, the objective of the current study was to estimate the longer term, legacy effect of randomization to intensive treatment on cardiovascular and all-cause mortality risk for participants in SPRINT. A secondary objective, in a subset of trial participants, was to examine attained SBP following the conclusion of trial follow-up. To accomplish these objectives, we linked SPRINT participants to the National Death Index (NDI) from 2016 to 2020, and extracted longitudinal outpatient measurements of SBP available in the electronic health record (EHR) over the same time period.

# METHODS

__Trial Design__: The design and methods of SPRINT have been published previously.[@ambrosius_design_2014; @sprint_research_group_randomized_2015] Briefly, it was a multicenter randomized clinical trial that compared two strategies for managing SBP in older adults with hypertension who were at increased risk of CVD. Participants were aged 50 years or older and had an SBP between 130 and 180 mm Hg at the screening visit, depending on the number of anti-hypertensive agents prescribed. Participants were considered to have an increased cardiovascular risk if they had clinical or subclinical cardiovascular disease, chronic kidney disease (CKD; defined by an estimated glomerular filtration rate of <60 mL/min/1.73 m^2^), or a Framingham Risk Score of 15% or greater or if they were aged 75 years or older. Individuals residing in a nursing home, persons with a diagnosis of dementia (based on medical record review), and those treated with medications primarily used for dementia therapy were excluded, as were persons with prevalent diabetes mellitus, history of stroke, proteinuria > 1 gram per day, or polycystic kidney disease. Individuals at 102 sites in the United States and Puerto Rico were randomized (1:1) to a SBP goal of less than 120 mm Hg (intensive treatment group, n = 4678) or a goal of less than 140 mm Hg (standard treatment group, n = 4683), using random permuted blocks with the randomization stratified by clinic site. The algorithms and formulary for the trial are listed in the published study protocol ( __citations needed__? ). Trial enrollment began in November 2010 and ended in March 2013, with active follow-up through July 1, 2016.[@sprintFinal] The study was approved by the institutional review board at each participating site, and each participant provided written informed consent. The study is registered at ClinicalTrials.gov (NCT01206062).

__Baseline Study Measurements__: Sociodemographic data were collected at baseline, with race or ethnicity information collected via self-report. The estimated glomerular filtration rate (eGFR) was calculated by the race-free 2021 CKD-EPI creatinine equation.[@inker_new_2021] Cognitive function was assessed using the Montreal Cognitive Assessment (MoCA) [@nasreddine_montreal_2005]. Lower cognitive function was defined as scoring at or below the estimated age and education-specific normative 10th percentile from the Irish Longitudinal Study of Aging,[@kenny_normative_2013] after adding 3 points to the scores of non-White participants.[@sachsMoCA2021] We defined frailty status at baseline using a 36-item Frailty Index (FI) based upon the model of deficit accumulation.[@pajewski_characterizing_2016] The FI is calculated as the sum of the score for each deficit divided by the total number of nonmissing items. We categorized frailty status as fit (FI ≤ 0.10), less fit (0.10 < FI ≤ 0.21), or frail (FI > 0.21).

__National Death Index Linkage__: Outcomes of interest included all-cause and CVD mortality. Methods of ascertainment and adjudication through the course of trial follow-up have been previously described.[@sprint_research_group_randomized_2015; @sprintFinal] Subsequently, mortality was ascertained through a US National Death Index (NDI) search. Possible matches were identified according to NDI guidelines.[@ndi_guide] Deaths were treated as confirmed if they were a Class 1 match, or a Class 2, 3, or 4 match with a probabilistic score above cutoffs recommended by the NDI.[@ndi_guide] NDI follow-up began in 2016 and ended on the date of death or December 31, 2020. Deaths ascertained in 2020 were based on the NDI preliminary data release. CVD mortality for NDI-based follow-up used the NDI Plus System, which automatically identifies underlying causes of death from death certificates, including conversion to ICD-10 codes. we defined CVD mortality as any death containing the ICD-10 codes of I00 to I99.[@regards_ndi]

__EHR Ancillary Study__: We examined the trajectory of SBP following the conclusion of the trial using outpatient SBPs extracted from the EHR. Methods for the linkage of participants to their medical record number and the extraction of vital sign data have been previously described.[@drawz_concordance_2020] Because encounter type information was inconsistently available (i.e. outpatient, inpatient, observation, etc.), we defined a BP measurement as outpatient if there were was not a BP measurement on the preceding or following day, and if there were 2 or less BP measurements on a particular day. We averaged outpatient EHR BP readings when there were 2 on the same day.

__Statistical Analysis__: Given the a priori expectation that treatment group differences may not be constant as a function of follow-up time (i.e. the proportional hazards assumption was likely to be invalid), we modeled treatment group differences as a function of time using two approaches. The first approach used Cox proportional hazards regression but split follow-up time into non-overlapping trial and cohort phases, estimating separate treatment group differences for each phase.[@zhang_time_varying_2018]. The second approach estimated the effect of treatment group as a continuous time-dependent effect.[@martinussen2007dynamic; @therneau2017using] All analyses accounted for correlation within study sites,[@lin_robust_1989] and analyses of cardiovascular mortality accounted for the competing risk of non-cardiovascular mortality.[@scheike_flexible_2008]. Mean between-group differences in outpatient SBP following the conclusion of trial follow-up were estimated using linear mixed models. Models included random effects for participant and clinic site and an interaction between treatment group and time since randomization, which was flexibly modeled using B-splines. All analyses were performed using SAS version 9.4 (SAS Institute Inc, Cary, NC) and R version `r paste(version$major, version$minor, sep = '.')` (R Project for Statistical Computing [http://www.r-project.org]) with multiple auxiliary R packages.[@r_language; @table.glue; @tidyverse; @survival_package; @timereg; @targets] All hypothesis tests were 2-sided, and P values less than 0.05 were considered statistically significant. No adjustments for multiple comparisons were made.

# RESULTS

___Study Participants___. A total of `r nrow(ndi_baseline)` participants were randomized between November 2010 and March 2013. The mean (standard deviation [SD]) age was `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years, with `r inline_text(tbl_characteristics, variable = 'age_cat', column = 'stat_0_1')`% of participants aged 75 years or older (__eTable 1__). Participants were `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% female and `r inline_text(tbl_characteristics, variable = 'race', column = 'stat_0_1')`% black. The mean (SD) SBP at baseline was `r inline_text(tbl_characteristics, variable = 'sbp', column = 'stat_0_1')` and `r inline_text(tbl_characteristics, variable = 'moca', column = 'stat_0_1')`% of participants MoCA scores were below the 10^th^ percentile. Participants included in the ancillary EHR study had greater age, lower SBP, less cognitive impairment, higher prevalence of CKD, and were more likely to be male or black than participants not included.

# `r inline_text(tbl_characteristics, variable = 'moca', column = 'stat_0_1')`

acm_overall_epochs <- fig |> 
  filter(variable == 'overall') |> 
  pull(eff_acm) |> 
  unlist(recursive = FALSE) |> 
  getElement('data') |> 
  filter(find_signchange_lgl(hr_upr-1)) |> 
  mutate(label = table_value(time))

t_signif <- acm_overall_epochs$label[1] 
t_atten <- acm_overall_epochs$label[2]

___All-cause Mortality___. In both treatment groups, median follow-up time was `r ap_b_o$fup_est` years. A total of `r ap_b_i$n_acm` and `r ap_b_s$n_acm` deaths occurred among participants randomized to intensive and standard treatment, respectively (__eTable 2__). The hazard ratio (HR) for all-cause mortality comparing intensive to standard treatment was `r with(ap_l_o_t, .("{mort_est} (95% confidence interval [CI] {mort_lwr}, {mort_upr})"))` during the trial phase, and `r with(ap_l_o_c, .("{mort_est} (95% CI {mort_lwr}, {mort_upr})"))` during the cohort phase. The continuous time-dependent effect of intensive versus standard treatment indicated a benefit for all-cause mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort phase (__Figure 1__).  In subgroups based on age, sex, race, CKD, cognitive function, and frailty, there was no evidence that intensive treatment during the trial phase had benefit for all-cause mortality during the cohort phase (__Figure 2__).

cvd_overall_epochs <- fig |> 
  filter(variable == 'overall') |> 
  pull(eff_cvd) |> 
  unlist(recursive = FALSE) |> 
  getElement('data') |> 
  slice(-1) |> 
  filter(find_signchange_lgl(hr_upr-1)) |> 
  mutate(label = table_value(time))

t_signif <- cvd_overall_epochs$label[1] 
t_atten  <- cvd_overall_epochs$label[2]

___Cardiovascular Mortality___. A total of `r ap_b_i$n_cvd` and `r ap_b_s$n_cvd` CVD mortality events occurred among participants randomized to intensive and standard treatment, respectively (__eTable 3__). The HR for CVD mortality among participants randomized to intensive versus standard treatment was `r with(ap_l_o_t, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the trial phase and `r with(ap_l_o_c, .("{cvd_est} (95% CI {cvd_lwr}, {cvd_upr})"))` during the cohort phase. The time-dependent effect of intensive versus standard treatment indicated a benefit for CVD mortality from `r t_signif` to `r t_atten` years from randomization, and was attenuated throughout the remainder of the cohort period (__Figure 3__). Adults randomized to intensive treatment who were <75 years of age, men, non-black, without CKD, or with cognitive function >10^th^ percentile had lower CVD mortality risk during the trial phase compared to their counterparts randomized to standard treatment, but there was no evidence that intensive treatment during the trial phase had benefit for CVD mortality during the cohort phase (__Figure 4__).

@@

# a temporary function for inline reporting below.
f <- function(d) with(d, .("{Estimate} ({Lower}, {Upper})"))

@@

t_atten <- bp_viz |> 
  getElement('diff') |> 
  getElement('data') |> 
  filter(source == 'EHR',
         find_signchange_lgl(Lower)) |> 
  pull(time_months) |> 
  divide_by(12)

@@

___Blood Pressure___. The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization (__Figure 5__). For participants randomized to standard treatment, mean (95% CI) SBP was estimated to be `r f(bp_viz_5yr$Standard)` at 5 years and `r f(bp_viz_10yr$Standard)` post-randomization. The difference in mean SBP levels between participants randomized to intensive versus standard treatment, in mm Hg, was `r f(bp_diff_5yr)` at 5 years post-randomization, and was attenuated at `r t_atten` years post-randomization (__eFigure 1__).

# DISCUSSION

A striking aspect of our results is the quickly weakening level of BP control for participants randomized to intensive treatment. While an attenuation of the between-group BP delta subsequent to the trial was certainly expected, one hypothesis was that such an attenuation would be driven by participants randomized to standard treatment pursuing a lower BP goal, given the results of SPRINT and subsequent changes to hypertension guidelines. However, this is clearly not what occurred. While we do not have access to prescription records to know how participant medication regimes may have changed after the trial, these results likely show some contribution of clinician therapeutic inertia, which has been identified as a significant barrier to improving population level control of hypertension. Combined with evidence showing recent decreases in the population of prevalence of controlled hypertension in the US,[@muntner2020htn] our results highlight the sobering reality facing the hypertension community. Sustainability is a clear limiting factor, especially with trying to implement lower BP goals and interventions earlier in adulthood.

This study has several limitations. First, while we restricted analyses to high quality NDI matches, some small degree of misclassification in linking participants to the NDI is likely. Second, while several studies have shown reasonable performance of using NDI diagnosis codes for defining CVD mortality, it is clearly not as robust as the adjudication process used in the primary follow-up for the trial. Third, information about BP control after the trial was limited to routine outpatient BPs extracted from the EHR, which are well known to poorly reflect the standardized BP measurement process used during the trial.[@drawz_concordance_2020]

# ACKNOWLEDGMENTS 
The views expressed in this paper are those of the authors and do not represent the official position of the National Institutes of Health (NIH), the National Heart, Lung, and Blood Institute, the Department of Veterans Affairs, or the U.S. Government, or the SPRINT Research Group. This paper was not reviewed by the SPRINT Publications and Presentations Committee. The authors also wish to acknowledge computing support provided the Veterans Affairs Informatics and Computing Infrastructure (VINCI).

# FINANCIAL DISCLOSURE 
The work presented here was supported by R01HL136679, R01AG055606, and R01AG065805. The Systolic Blood Pressure Intervention Trial was funded by the National Institutes of Health (including the National Heart, Lung, and Blood Institute, the National Institute of Diabetes and Digestive and Kidney Diseases, the National Institute on Aging, and the National Institute of Neurological Disorders and Stroke) under contracts HHSN268200900040C, HHSN268200900046C, HHSN268200900047C, HHSN268200900048C, and HHSN268200900049C and interagency agreement A-HL-13-002-001. It was also supported in part with resources and use of facilities through the Department of Veterans Affairs. Azilsartan and chlorthalidone (combined with azilsartan) were provided by Takeda Pharmaceuticals International Inc.

Additional support was provided through the following National Center for Advancing Translational Sciences clinical and translational science awards: UL1TR000439 (awarded to Case Western Reserve University); UL1RR025755 (Ohio State University); UL1RR024134 and UL1TR000003 (University of Pennsylvania); UL1RR025771 (Boston University); UL1TR000093 (Stanford University); UL1RR025752, UL1TR000073, and UL1TR001064 (Tufts University); UL1TR000050 (University of Illinois); UL1TR000005 (University of Pittsburgh); U54TR000017-06 (University of Texas Southwestern Medical Center); UL1TR000105-05 (University of Utah); UL1 TR000445 (Vanderbilt University); UL1TR000075 (George Washington University); UL1 TR000002 (University of California, Davis); UL1 TR000064 (University of Florida); and UL1TR000433 (University of Michigan); and by National Institute of General Medical Sciences, Centers of Biomedical Research Excellence award NIGMS P30GM103337 (awarded to Tulane University).

```{r tbl_footers}

ckd_defn <- as_paragraph("Chronic Kidney Disease was defined by an estimated glomerular filtration rate <60 ml/min/1.73 m", as_sup(2L), " based on the 2021 CKD-EPI creatinine equation.")

```{r tbl_abbrvs}

abbrvs <- list(
  CI   = 'confidence interval',
  FI   = 'frailty index',
  SD   = 'standard deviation',
  IQR  = 'interquartile range',
  MoCA = 'Montreal cognitive assessment',
  EHR  = 'electronic health records',
  CVD  = 'cardiovascular disease'
)

abbrvs_paste <- function(strings){

  front <- 'Abbreviations'

  matter <- glue::glue_collapse(strings, sep = '; ', last = '; and ')

  paste(front, matter, sep = ': ')

}

abbrvs_write <- function(abbr){

  sorted <- abbr[sort(names(abbr))]

  strings <- map2_chr(names(sorted), sorted, paste, sep = ' = ')

  as_paragraph(abbrvs_paste(strings))

}

<!-- new page not needed at the end of a block section -->
```{r}

margins_normal <- page_mar(bottom = 2.54 * 0.3937, 
                           top    = 2.54 * 0.3937, 
                           right  = 2.54 * 0.3937, 
                           left   = 2.54 * 0.3937)

block_section(
  prop_section(
    page_size = page_size(orient = "portrait"),
    type = "continuous",
    page_margins = margins_normal
  )
)

__Figure 1__: (A) Cumulative incidence of all-cause mortality by treatment group. (B) Time-dependent effect of randomization to intensive treatment for all-cause mortality. `r single_par`

@@

plot_grid(
  # row 1 (legend)
  legend, 
  blank_plot,
  # row 2 (figures)
  fig$inc_acm[[1]]$plot + theme(legend.position = ''), 
  fig$eff_acm[[1]],
  # row 3 (table)
  fig$inc_acm[[1]]$table, 
  blank_plot,
  ncol = 2,
  align = 'v',
  axis = 'lr',
  rel_heights = c(1,10,2),
  labels = c("","","A","B","",""),
  label_x = 0.1,
  label_y = 1.05
)

\newpage
__Figure 2__: All-cause mortality hazard ratio for participants randomized to intensive versus standard treatment.

long_viz_mort

\newpage
__Figure 3__: (A) Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group. (B) Time-dependent effect of randomization to intensive treatment for cardiovascular mortality. `r single_par`

@@

\newpage
__Figure 4__: Cardiovascular mortality hazard ratio for participants randomized to intensive versus standard treatment.

long_viz_cvd

\newpage
__Figure 5__: Mean systolic blood pressure over time by treatment group. Shaded areas indicate a 95% confidence interval for the mean. `r single_par`

bp_viz$levels + 
  theme(legend.key.size = unit(1, 'cm'))

# fpath <- file.path(
#   paste0(get_sasinet_drive(), ":"),
#   'bjaeger',
#   'sprint-mortality-ndi',
#   'fig',
#   'SPRINT_EHR_SBP_LongTerm_Comparison_122721.png'
# )
# 
# knitr::include_graphics(
#   path = fpath, 
#   dpi = 600
# )

```{r}
block_section(
  prop_section(
    page_size = page_size(orient = "landscape"),
    page_margins = page_mar(bottom = 1/8, top = 1/8, 
                            right = 1/4, left = 1/4),
    type = "continuous"
  )
)
```

__Online supplement only__

Byron C. Jaeger, PhD,^1^ Adam P. Bress, PharmD, MS,^2,3^ Joshua D. Bundy, PhD,^4^ Alfred K. Cheung, MD,^5,6^ William C. Cushman, MD,^7^ Paul E. Drawz, MD,^8^ Lawrence J. Fine, MD,^9^ Karen C. Johnson, MD,^7^ Cora E. Lewis, MD,^10^ Suzanne Oparil, MD,^11^ Michael V. Rocco, MD,^12^ Stephen R. Rapp, PhD, ^13^ Joni K. Snyder, RN,BSN,^9^ Mark A. Supiano, MD,^14^ Paul K. Whelton, MD,^4^ Jeff D. Williamson, MD, MHS,^15^ Jackson T. Wright, Jr, MD, PhD,^16^ David M. Reboussin, PhD,^1^ and Nicholas M. Pajewski, PhD^1^

^1^Department of Biostatistics and Data Science, Wake Forest School of Medicine, Winston-Salem, NC. ^2^Informatics, Decision-Enhancement, and Analytic Sciences (IDEAS) Center, Veterans Affairs, Salt Lake City Health Care System, Salt Lake City, UT. ^3^Department of Population Health Sciences, University of Utah School of Medicine, Salt Lake City, UT. ^4^Department of Epidemiology, Tulane University School of Public Health and Tropical Medicine, New Orleans, LA. ^5^Renal Section, Veterans Affairs Salt Lake City Healthcare System, UT. ^6^Division of Nephrology and Hypertension, Department of Internal Medicine, University of Utah School of Medicine, Salt Lake Cite, UT. ^7^Department of Preventive Medicine, University of Tennessee Health Science Center, Memphis. ^8^Division of Renal Diseases & Hypertension, University of Minnesota, MN. ^9^Clinical Applications and Prevention Branch, National Heart, Lung, and Blood Institute, National Institutes of Health, Bethesda, MD. ^10^Department of Epidemiology, University of Alabama at Birmingham. ^11^Division of Cardiovascular Disease, Department of Medicine, University of Alabama at Birmingham. ^12^Section on Nephrology, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC. ^13^Departments of Psychiatry and Behavioral Medicine & Social Sciences and Health Policy, Wakfe Forest School of Medicine, Winston-Salem, NC. ^14^Division of Geriatrics, University of Utah School of Medicine, Salt Lake City. ^15^Section on Gerontology and Geriatric Medicine, Department of Internal Medicine, Wake Forest School of Medicine, Winston-Salem, NC. ^16^Division of Nephrology and Hypertension, University Hospitals Cleveland Medical Center, Case Western Reserve University, OH `r single_par`
\

```{r}
block_section(
  prop_section(
    page_size = page_size(orient = "portrait"),
    type = "continuous",
    page_margins = margins_normal
  )
)
```

__eTable 1__: Characteristics of study participants.`r single_par`

header_labels <- tbl_characteristics |> 
  show_header_names(quiet = TRUE) |> 
  mutate(
    label = str_remove_all(label, fixed("*")),
    label = if_else(column == 'stat_0_1', 
                    paste0("Overall, ", label),
                    label),
    label = str_replace(label, ', ', '\n')
  ) |> 
  deframe()

tbl_characteristics |> 
  modify_footnote(update = everything() ~ NA) |> 
  modify_spanning_header(update = everything() ~ NA) |> 
  as_flex_table() |> 
  set_header_labels(
    label = "Characteristic",
    stat_0_1 = "Overall\nN = 9,361",
    stat_1_2 = "Standard\nN = 4,683",
    stat_2_2 = "Intensive\nN = 4,678",
    p.value_2 = "p-value",
    stat_1_3 = "No\nN = 5,952",
    stat_2_3 = "Yes\nN = 3,409",
    p.value_3 = "p-value"
  ) |> 
  add_header_row(
    values = c("Characteristic",
               "Overall\nN = 9,361",
               "Treatment",
               header_labels['p.value_2'],
               "Included in EHR ancillary study", 
               header_labels['p.value_3']),
    colwidths = c(1, 1, 2, 1, 2, 1)
  ) |> 
  theme_box() |> 
  merge_v(part = 'header') |> 
  align(align = 'center', part = 'all') |> 
  align(j = 1, align = 'left', part = 'all') |> 
  footnote(
    i = 1, 
    j = 1, 
    ref_symbols = '',
    value = as_paragraph(
      "Table values are percentages unless otherwise specified"
    )
  ) |> 
  footnote(
    i = 1, 
    j = ~p.value_2 + p.value_3,
    part = 'header',
    ref_symbols = 'a',
    value = as_paragraph(
      "P-values computed using Wilcoxon rank sum test for age in years, chi-square test for age \u226575 years, and chi-square test for all other variables"
    )
  ) |> 
  footnote(
    i = ~ label == "Chronic Kidney Disease",
    j = 1, 
    ref_symbols = 'b',
    value = ckd_defn,
    part = 'body'
  ) |> 
  footnote(
    i = 1,
    j = 1,
    ref_symbols = '',
    value = abbrvs_write(abbrvs[c('SD', 'EHR', 'IQR', 'MoCA', 'FI')]),
    part = 'body'
  ) |>
  fontsize(size = 12, part = 'all') |> 
  font(fontname = 'Times New Roman', part = 'all') |> 
  width(j = ~ stat_0_1 + stat_1_2 + stat_2_2 + stat_1_3 + stat_2_3,
        width = 1) |> 
  width(j = 1, width = 1.75)

__eTable 2__: All-cause mortality by treatment group and subgroup `r single_par`

tbl_inf <- long_inf |> 
  mutate(mort = .("{mort_est}\n({mort_lwr}, {mort_upr})")) |> 
  select(variable,
         group, 
         trial_phase,
         mort) |> 
  pivot_wider(names_from = trial_phase,
              values_from = mort,
              names_prefix = '.h_')

tbl_data <- long_ds |>
  mutate(
    across(starts_with("n_"), as.integer),
    .n = .("{n_acm} / {n_obs}"),
    .i = .("{inc_mort_est}\n({inc_mort_lwr}, {inc_mort_upr})")
  ) |>
  select(-starts_with("inc_"),
         -starts_with("n_")) |> 
  pivot_wider(names_from = c(treatment, trial_phase),
              values_from = c(.n, .i)) |> 
  left_join(tbl_inf) |> 
  mutate(
    variable = replace(variable, group == variable, NA_character_),
    variable = recode(variable, !!!recoders$variable),
    group = recode(group, !!!recoders$group)
  )

>>>>

---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}

knitr::opts_chunk$set(echo = FALSE, 
                      collapse = TRUE,
                      comment = "#>",
                      message = FALSE,
                      warning = FALSE,
                      dpi = 600, 
                      fig.width = 11, 
                      fig.height = 6.25)

# specification for tables
rspec <- round_spec() |>
  round_using_magnitude(digits = c(2, 1, 0),
                        breaks = c(10, 100, Inf))

@@

@@

# sprint-mortality-ndi

<!-- badges: start -->
<!-- badges: end -->

@@

@@

# a temporary function for inline reporting of BP values
f <- function(d) with(d, .("{Estimate} ({Lower}, {Upper})"))

@@

t_atten_sbp <- bp_viz |> 
  getElement('diff') |> 
  getElement('data') |> 
  filter(source == 'EHR',
         find_signchange_lgl(Lower)) |> 
  pull(time_months) |> 
  divide_by(12)

@@

__IMPORTANCE__. The Systolic Blood Pressure Intervention Trial (SPRINT) showed that intensive treatment, defined by a systolic blood pressure (SBP) goal of <120 mmHg, reduced the risk of cardiovascular morbidity and mortality. However, the post-trial association of intensive treatment with mortality is unknown.

__OBJECTIVE__. To evaluate the association of receiving intensive treatment for approximately three years during the SPRINT trial with all-cause and cardiovascular mortality up to ten years post-randomization.

__DESIGN, SETTING, AND PARTICIPANTS__. SPRINT, a randomized clinical trial of 9361 patients aged 50 years or older with hypertension and increased cardiovascular risk, but without diabetes or history of stroke. Randomization began on November 8, 2010, the trial intervention ended on August 20, 2015, and trial follow-up visits occurred through July 2016.

__INTERVENTIONS__. Randomization to SBP goal of either <120 mmHg (intensive treatment, N=4678) versus <140 mmHg (standard treatment, N=4683).

__MAIN OUTCOMES AND MEASURES__. All-cause and cardiovascular mortality assessed via the US National Death Index, beginning in 2016 through December 31st, 2020. In a subset of 3644 trial participants, outpatient SBP levels measured in routine clinical practice after the trial were examined.

__RESULTS__. Among 9361 randomized participants (mean [standard deviation] age `r inline_text(tbl_characteristics, variable = 'age_yrs', column = 'stat_0_1')` years; `r inline_text(tbl_characteristics, variable = 'sex', column = 'stat_0_1')`% women), the median intervention phase was 3.3 years. Over a median follow-up of `r ap_b_o$fup_est` years, intensive treatment was beneficial for both cardiovascular (Hazard Ratio [HR] = `r with(ap_l_o_t, .("{cvd_est}, 95% confidence interval [CI] {cvd_lwr} to {cvd_upr}"))`) and all-cause mortality (HR = `r with(ap_l_o_t, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`) through close-out visits for the trial. However, there was no evidence of association during post-trial follow-up for cardiovascular (HR = `r with(ap_l_o_c, .("{cvd_est}, 95% CI {cvd_lwr} to {cvd_upr}"))`) or all-cause mortality (HR = `r with(ap_l_o_c, .("{mort_est}, 95% CI {mort_lwr} to {mort_upr}"))`). The estimated mean (95% CI) SBP among participants randomized to intensive treatment was `r f(bp_viz_5yr$Intensive)` at 5 years and `r f(bp_viz_10yr$Intensive)` at 10 years post-randomization.

__CONCLUSIONS AND RELEVANCE__. The beneficial effect of intensive treatment on cardiovascular and all-cause mortality was attenuated during post-trial observational follow-up. Given increasing SBP levels in participants randomized to intensive treatment following the trial, these results highlight the importance of consistent long-term management of hypertension.

__CENTRAL ILLUSTRATION__

Panel A: Cumulative incidence of cardiovascular and non-cardiovascular mortality by treatment group. Panel B: Time-dependent effect of randomization to intensive treatment for cardiovascular mortality.

@@

@@
